Reasoning Graphically about Quantum Computation

### 2008-03-02 papers/reasoning-graphically/paper.tex
working...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1391 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\noindent This motivates a representation that has only one
representative for such equations. Such a representation abstracts
over the boundary nodes membership in the domain and co-domain. In
this section we formalise such a notion and refer to the generalised
boundary nodes as variable-nodes.

The intuition of a variable node is that it can be replaced by any
subgraph of the right type. Thus, unlike nodes in conrete a graph,
variable nodes do not define an ordering on the edges that are
incident. Instead, the edges that touch a variable node are simply a
set. The type of a variable nodes is thus defined as a multiset rather
than a tensor. Any tensor type can be lifted to a multiset type in the
obvious manner, which we denote by the operation {\tt
  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$ is
an instance of a multiset type $M$, written $T \in M$, when $M
\subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
when a variable node is instantiated to some other node, it's type
will have at least the currently attached edges.

A graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs which is denoted by
$\vinterp{G}$. We refer to members of this set as instances. An
instance of a graph with variable nodes is the graph with every
variable node replaced by a subgraph of the right type. When a graph
has several varibale-nodes, the replacement subgraphs may be
intersecting. For example, in
Figure~\ref{node-variable-instances-fig}, the graphs $a$, $b$ and $c$
are instances of the graph $x$. A graph with variable nodes, and
without boundary nodes, is called a \emph{variable-node graph}. Such
graphs are complete in the sense that for every concrete graph there
is a variable node graph that can be instantiated to it: $\forall
g.\,\exists G.\,g \in \vinterp{G}$. The witness for this proof is made
by simply replacing all bounary nodes with variable nodes.

@@

@@

@@

A crucial feature of rules like the Spider Law is that an arbirary
number of repetitions of a subgraph to be matched. To support this we
introduce an operation, !-boxing, on graph representations. Given a
graph representation, this introduces a new notation by outlining
(called !-boxing) a set of chosen nodes. Intuitively, the resulting
!-boxed graph can be thought of as representing a set of graphs with
an arbirary number of copies of the !-boxed nodes, where every copy
has the connects to the same nodes outside the !-box. The !-boxes can
also be instantiated with zero copies which is erases all edges from
the !-box. See Figure~\ref{bang-box-ex1-fig} for a simple example.

More formaly, a graph with !-boxes is simply a graph paired with a set
!-boxes. Each !-box is just the set of nodes inside it and each !-box
must be disjoint from the others.\footnote{The disjointness condition
  avoids a slightly more complex, although also more expressive,
  semantics for overlaping node sets in the !-boxes. While such
  expressivity is interesting from a representational point of view,
  it is not needed to express the currently known rules of quantum
  computation.}

@@

The set of !-box graphs matched by a !-box graph is defined as the set
closed by the following operations:

\begin{definition}
  \emph{copy} copies a !-box in a graph to produce a new graph with
  two copies of the boxed subgraph, with a second box around the newly
  copied nodes. Any edges between a node inside the !-box, $n$ and one
  outside it $m$ get copied so that there is a new edge from $m$ to
  the new copy of $n$.
\end{definition}

\begin{definition}
  \emph{drop} simply removes the box, but leaves its contents in the
  graph.
\end{definition}

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box set as
  well as any incident edges to those nodes.
\end{definition}

@@

\noindent We use $G \leq_! H$ to denote that $G$ matches $H$.

@@

We give a formal semantics to !-boxing a representation of graphs by
defining the !-box graph in terms of a set of graphs in the
underlying representation. We denote the interpretation of a graph $G$
by $\binterp{G}$ and say that mebemrs of $\binterp{G}$ are instances
of $G$.

@@

The purpose of graph patterns is to formally express, in a declarative
manner, the intuitive idea of graph matching commonly used in for
Quantum Computation. To do this, we combine the two extensions to the
concrete graph representation, namely variable nodes and !-boxes, to
provides a matching relation between graph patterns and concrete
graphs as well as directly between graph patterns. The former relation
gives a semantics for our system and ensures soundness, while the
latter supports deirved equations and thus provides the theoretical
basis for our formulation of rewriting.

@@

The semantics for a graph pattern $G$ is a set of concrete graphs
denoted by $\minterp{G}$. We say that members of this set are
\emph{matched} by $G$. Formally, the set of matches are defined as:

@@

In this section we outline how the formalism we have presented for
graph patterns can be used to provide a \emph{meta-level} framework
for reasoning about a given model of a compact closed
categories. Following the terminology of logical frameworks such as
Isabelle~\cite{isabelle}, we call the specification provided by the
underlying model an \emph{object-level} graph formalism. Each
object-level formalism defines a set of rules which are treated as the
axioms for the system. The object-level implementation also defines
the data at the nodes and edges as well as corresponding data-matching
behaviour.

\subsection{Substitution with Equational Rules}

A rule in our framework declarative in the sense that it is a pair of
pattern graph that represent the left and right hand sides of an
equation. Substituon with a rule then involves replacing a subgraph
that matches the left hand side with the right hand side. To do this
there has to be a isomorphism between variable nodes in the left and
right hand subgraphs. More concretely, there must be a unique naming
of the variable nodes in the left hand side such that each name occurs
in a unique variable node on the right hand side. Given a matching on
the left hand side, the target subgraph can be replaced with right
hand side connected by the corresponding variable nodes. This provides
rewriting for pattern graphs.

@@

\item The soundness and adequacy of the axioms when expressed within
  the compact closed category. Adquency ensures that the encoding as a
  graph pattern represents only the intended rules.  The soundness of
  the rules is obviously needed and requires proving the rule in terms
  of the underlying semantics.
\end{itemize}

The completeness of the resulting system, with respect to the intended
semantics, depends on the completeness of the set of given rules. The
completeness of

>>>>

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we need a
representation that abstracts over the boundary nodes membership in
the domain or co-domain. In this section we formalise such a notion
and refer to the generalised boundary nodes as \emph{variable
  nodes}. This gives rise to a \emph{variable-node graphs}, which are
graphs with variable nodes instead of boundary edges.

@@

\subsection{Semantics}

@@

The intuition of variable nodes is that they can replaced by any be
nodes of the right type that are themseleves within any graph.  A
graph with variable nodes, $G$, can be given a formal semantics by
being interpretated as a set of concrete graphs, denoted by
$\vinterp{G}$.

% Thus, unlike nodes in conrete a graph, variable nodes do not define an
% ordering on the edges that are incident. Instead, the edges that touch
% a variable node are simply a set. The type of a variable nodes is thus
% defined as a multiset rather than a tensor. Any tensor type can be
% lifted to a multiset type in the obvious manner, which we denote by
% the operation {\tt mset\_of\_tensor}. For example, {\tt
%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
% say that a tensor-type $T$ is an instance of a multiset type $M$,
% written $T \in M$, when $M \subseteq
% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
% variable node is instantiated to some other node, it's type will have
% at least the currently attached edges.

% An instance of a graph with variable nodes is the graph with every
% variable node replaced by a subgraph of the right type. When a graph
% has several varibale-nodes, the replacement subgraphs may be
% intersecting.

We refer to members of this set as instances.

For example, in Figure~\ref{node-variable-instances-fig}, the graphs
$a$, $b$ and $c$ are instances of the graph $x$. A graph with variable
nodes, and without boundary nodes, is called a \emph{variable-node
  graph}.

\subsubsection{Completness and Adequacy.}

Variable-node graphs are complete in the sense that for every concrete
graph there is a variable node graph that can be instantiated to it:
$\forall g.\,\exists G.\,g \in \vinterp{G}$. The witness for this
proof is made by simply replacing all bounary nodes with variable
nodes.

@@

A crucial feature of rules like the Spider Law is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing, on graph
representations. Given a graph representation, this introduces a new
notation by outlining (called !-boxing) a set of chosen
nodes. Intuitively, the resulting !-boxed graph can be thought of as
representing a set of graphs with an arbirary number of copies of the
!-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies, which erases all edges to and from the !-box. See
Figure~\ref{spider-bang-rule} .

More formaly, a graph with !-boxes is a graph paired with a set
!-boxes. Each of these !-boxes is the set of nodes inside it and each
!-box must be disjoint from the others.\footnote{The disjointness
  condition avoids a slightly more complex, although also more
  expressive, semantics for overlaping node sets in the !-boxes. While
  such expressivity is interesting from a representational point of
  view, it is not needed to express the currently known rules of
  quantum computation.}

@@

To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from it's nodes, we introduce
\emph{!-box matching}. This is a binary relation on graphs, written
infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
matched by a !-box graph is defined as the set closed by the following
operations on the !-boxes in a graph:

\begin{definition}
  \emph{copy}: copies a !-box, $b$ in a graph to produce a new graph
  with two copies of the !-boxed subgraph, $b$ is the old one and $b'$
  is the new one. The set of !-boxes in the copied graph now also
  contains the new !-box $b'$. Any edges between a node, $n$, inside
  the !-box $b$, and a node, $m$, outside it, get copied so that there
  is a new edge from $m$ to the new copy of $n$ in $b'$.
\end{definition}

\begin{definition}
  \emph{drop} simply removes the !-box, but leaves its contents in the
  graph.
\end{definition}

\begin{definition}
  \emph{kill} removes from the graph all nodes in the !-box as well as
  any incident edges.
\end{definition}

@@

@@

We give a formal semantics to a !-box graphs by defining the !-box
graph in terms of a set of graphs in the underlying representation. We
denote the interpretation of a !-box graph $G$ by $\binterp{G}$ and
say that members of $\binterp{G}$ are instances of $G$.

@@

The purpose of graph patterns is to formally express, in a declarative
manner, the intuitive idea of graph matching which commonly used in
quantum computation. To do this, we combine the two extensions to the
concrete graph representation, namely variable nodes and !-boxes, to
provides a matching relation between graph patterns and concrete
graphs as well as directly between graph patterns. The former relation
gives a semantics for our system and ensures soundness, while the
latter supports deirved equations and thus provides the theoretical
basis for our formulation of rewriting.

\subsubsection{Completeness.}

@@

The semantics for a graph pattern $G$ is a set of concrete graphs
denoted by $\minterp{G}$. We say that members of this set are
\emph{matched} by $G$. Formally, the concrete set of matches are
defined as:

@@

In this section we outline how the formalism we have presented for
graph patterns can be used to provide a \emph{meta-level} framework
for reasoning about models of compact closed categories. Following the
terminology of logical frameworks such as Isabelle~\cite{isabelle}, we
call the specification provided by the underlying model an
\emph{object-level} graph formalism.

An object-level formalisation defines a set of rules which are treated
as the axioms for the system. The object-level implementation also
defines the data at the nodes and edges as well as corresponding
data-matching behaviour. The meta-level provides machinery to
manipulate graphs and derive new equations. In this section we
describe the meta-level framework and prove its adequacy for
rewriting.

\subsection{Equational Rules}

Axioms defined by an object-level model, as well as derived rules in
our framework, are declarative in the sense that they are pairs of
pattern graph. This pair represents the left and right hand sides of
an equation.

Substituon with a rule then involves replacing a subgraph that matches
the left hand side with the right hand side. To do this there has to
be a isomorphism between variable nodes in the left and right hand
subgraphs. More concretely, there must be a unique naming of the
variable nodes in the left hand side such that each name occurs in a
unique variable node on the right hand side. Given a matching on the
left hand side, the target subgraph can be replaced with right hand
side connected by the corresponding variable nodes. This provides
rewriting for pattern graphs.

\subsubsection{Adequacy}

A an equation that involves a graph pattern on one side

\emph{Adequacy} says that if one by doing so within an
equation, we are not

@@

\item The soundness and adequacy of the axioms when expressed within
  the compact closed category. Adquency ensures that the encoding as a
  graph pattern represents only the intended concrete rules. The
  soundness of the rules is obviously needed and requires proving the
  rule in terms of the underlying semantics.
\end{itemize}

The proof of adequancy ...

The completeness of the resulting system, with respect to the intended
underlying semantics, depends on the completeness of the set of given
rules.
